---
title: 'RBC 16S analysis: abiotic variables'
output: html_document
---

Written by Delaney Beals \| Date created: October 26, 2023 \| Date most recent edit: December 11, 2023

Resource: <https://rstudio-pubs-static.s3.amazonaws.com/548738_f2de141dddb146879c535583eb8ba79c.html>

Here we want to determine if there is a relationship between different environmental parameters that don't involve the microbial communities of the soil. We are mostly interested in the relationship between methane flux and other variables. Data was obtained from the Knowlton Fork (KF) 2021 datasheets available on HydroShare.

```{r message=FALSE, warning=FALSE}
library(vegan)
library(tidyverse)
library(ggplot2)
library(viridis)
library(cowplot)
library(lubridate)
library(ggpubr)
library(corrplot)
library(extrafont)
```

```{r}
# Load in data
abiotic_metadata <- read.csv("data/KF_metadata.csv", header = T, row.names = 1, sep = ",")
abiotic <- as.data.frame(abiotic_metadata)

head(abiotic)
```

Next we will run Pearson correlation tests between different variables.

```{r}
# flux vs. water dissolved oxygen
cor.test(abiotic$flux.med, abiotic$dissolved.oxg.mgL, 
                    method = "pearson") # R = 0.8871995; p-value = 0.0447
cor.test(abiotic$flux.med.rip, abiotic$dissolved.oxg.mgL, 
                    method = "pearson") # R = 0.9104768; p-value = 0.03172
cor.test(abiotic$flux.med.up, abiotic$dissolved.oxg.mgL, 
                    method = "pearson") # R = 0.8033748; p-value = 0.1015

# flux vs. plant cover
cor.test(abiotic$flux.med, abiotic$KF.plant.cover, 
                    method = "pearson") # R = -0.4950569; p-value = 0.3965
cor.test(abiotic$flux.med.rip, abiotic$KF.plant.cover, 
                    method = "pearson") # -0.5904275 p-value = 0.2946
cor.test(abiotic$flux.med.up, abiotic$KF.plant.cover, 
                    method = "pearson") # -0.3597735  p-value = 0.552

# flux vs. water temperature
cor.test(abiotic$flux.med, abiotic$WaterTemp_EXO, 
                    method = "pearson") # R = -0.9586511; p-value = 0.01003
cor.test(abiotic$flux.med.rip, abiotic$WaterTemp_EXO, 
                    method = "pearson") # R = -0.9304217; p-value = 0.0218
cor.test(abiotic$flux.med.up, abiotic$WaterTemp_EXO, 
                    method = "pearson") # R = -0.9119055; p-value = 0.03097

# flux vs. volumetric water content at 5 cm
cor.test(abiotic$flux.med, abiotic$KF.VWC, 
                    method = "pearson") # R = 0.7352166; p-value = 0.1569
cor.test(abiotic$flux.med.rip, abiotic$KF.VWC, 
                    method = "pearson") # R = 0.6397685; p-value = 0.245
cor.test(abiotic$flux.med.up, abiotic$KF.VWC, 
                    method = "pearson") # R = 0.741945;  p-value = 0.1511

# flux vs. air temperature
cor.test(abiotic$flux.med, abiotic$KF.air.temp, 
                    method = "pearson") # R = -0.5135805; p-value = 0.3761
cor.test(abiotic$flux.med.rip, abiotic$KF.air.temp, 
                    method = "pearson") # R = -0.616543;  p-value = 0.268
cor.test(abiotic$flux.med.up, abiotic$KF.air.temp, 
                    method = "pearson") # R = -0.3708015; p-value = 0.5389

# flux vs. 5 cm soil temperature
cor.test(abiotic$flux.med, abiotic$KF.soil.temp, 
                    method = "pearson") # R = -0.959627; p-value = 0.009679
cor.test(abiotic$flux.med.rip, abiotic$KF.soil.temp, 
                    method = "pearson") # R = -0.9522546; p-value = 0.01243
cor.test(abiotic$flux.med.up, abiotic$KF.soil.temp, 
                    method = "pearson") # R = -0.9008488; p-value = 0.03692

# flux vs. gravimetric water content
cor.test(abiotic$flux.med.up, abiotic$GWC.up, 
                    method = "pearson") # R = -0.3526934;  p-value = 0.5604
cor.test(abiotic$flux.med.rip, abiotic$GWC.rip, 
                    method = "pearson") # R = -0.6789623 ; p-value = 0.2075

# flux vs. permittivity
cor.test(abiotic$flux.med, abiotic$permit, 
                    method = "pearson") # R = 0.7871688; p-value = 0.114
cor.test(abiotic$flux.med.rip, abiotic$permit, 
                    method = "pearson") # R = 0.8002092; p-value = 0.1039
cor.test(abiotic$flux.med.up, abiotic$permit, 
                    method = "pearson") # R = 0.694907; p-value = 0.1928

# flux vs. wind speed
cor.test(abiotic$flux.med, abiotic$wind, 
                    method = "pearson") # R = -0.2688901; p-value = 0.6618
cor.test(abiotic$flux.med.rip, abiotic$wind, 
                    method = "pearson") # R = -0.1995695; p-value = 0.7476
cor.test(abiotic$flux.med.up, abiotic$wind, 
                    method = "pearson") # R = 0.167129; p-value = 0.7882

# flux vs. water pH
cor.test(abiotic$flux.med, abiotic$pH, 
                    method = "pearson") # R = -0.2235258; p-value = 0.7178
cor.test(abiotic$flux.med.rip, abiotic$pH, 
                    method.up = "pearson") # R = -0.1370048; p-value = 0.8261
cor.test(abiotic$flux.med.up, abiotic$pH, 
                    method = "pearson") # R = -0.3815217 ; p-value = 0.5263

# flux vs. dew point
cor.test(abiotic$flux.med, abiotic$dew, 
                    method = "pearson") # R = -0.5296249; p-value = 0.3587
cor.test(abiotic$flux.med.rip, abiotic$dew, 
                    method = "pearson") # R = -0.5904697; p-value = 0.2945
cor.test(abiotic$flux.med.up, abiotic$dew, 
                    method = "pearson") # R = -0.4023638; p-value = 0.5019

# flux vs. relative humidity
cor.test(abiotic$flux.med, abiotic$RH, 
                    method = "pearson") # R = -0.5413367; p-value = 0.3461
cor.test(abiotic$flux.med.rip, abiotic$RH, 
                    method = "pearson") # R = -0.4708523; p-value = 0.4234
cor.test(abiotic$flux.med.up, abiotic$RH, 
                    method = "pearson") # R = -0.521966; p-value = 0.367

# flux vs. VWC at 10 cm
cor.test(abiotic$flux.med, abiotic$VWC.10, 
                    method = "pearson") # R = 0.7422863 ; p-value = 0.1508
cor.test(abiotic$flux.med.rip, abiotic$VWC.10, 
                    method = "pearson") # R = 0.7489831 ; p-value = 0.1452
cor.test(abiotic$flux.med.up, abiotic$VWC.10, 
                    method = "pearson") # R = 0.6456267 ;  p-value = 0.2393

# flux vs. VWC at 20 cm
cor.test(abiotic$flux.med, abiotic$VWC.20, 
                    method = "pearson") # R = 0.8656571  ; p-value = 0.0579
cor.test(abiotic$flux.med.rip, abiotic$VWC.20, 
                    method = "pearson") # R = 0.860721  ; p-value = 0.06108
cor.test(abiotic$flux.med.up, abiotic$VWC.20, 
                    method = "pearson") # R = 0.7809968  ;  p-value = 0.1189

# flux vs. VWC at 50 cm
cor.test(abiotic$flux.med, abiotic$VWC.50, 
                    method = "pearson") # R = 0.6229855   ; p-value = 0.2616
cor.test(abiotic$flux.med.rip, abiotic$VWC.50, 
                    method = "pearson") # R = 0.6594483   ; p-value = 0.226
cor.test(abiotic$flux.med.up, abiotic$VWC.50, 
                    method = "pearson") # R = 0.4978925   ;  p-value = 0.3933

# flux vs. VWC at 100 cm
cor.test(abiotic$flux.med, abiotic$VWC.100, 
                    method = "pearson") # R = -0.5978083    ; p-value = 0.287
cor.test(abiotic$flux.med.rip, abiotic$VWC.100, 
                    method = "pearson") # R = -0.4667811    ; p-value = 0.428
cor.test(abiotic$flux.med.up, abiotic$VWC.100, 
                    method = "pearson") # R = -0.7157318   ;  p-value = 0.174
```

```{r}
abiotic1 <- abiotic[,-26]
abiotic2 <- abiotic1[,-1]
#calculate correlation
corr.abio <- Hmisc::rcorr(as.matrix(abiotic2), type=c("pearson"))

# plot correlation
corrplot(corr=corr.abio$r[2:15, 16:28], 
         p.mat=corr.abio$P[2:15, 16:28], 
         tl.col = "black",
         tl.srt = 45,
         insig = "pch", 
         sig.level = c(.05), 
         pch.cex = 2, 
         pch.col = "black", 
         number.cex=0.7,
         number.digits = 2,
         cl.length = 5,
         cl.align.text = "l")

# to get P values
res8 <- cor.mtest(abiotic2, conf.level = .95)

# another way to generate plot
corrplot(cor(abiotic2, method = "pearson"), tl.col = "black", tl.cex = 0.7, p.mat = res8$p, insig = "pch", sig.level = c(.05), pch.cex = 1, pch.col = "black",  order = "hclust", hclust.method = "average")
```

We can also plot these relationships. The geom_smooth function will add a linear regression trend line over our data

```{r}
# for editing theme
my_theme <- theme_bw() +
        theme(text=element_text(size = 12, family = "Calibri"),
        axis.text.x=element_text(size=12, color = "black", family = "Calibri"), 
        axis.text.y = element_text(size=12, color = "black",  family = "Calibri"),
        title = element_text(size = 12, color = "black",  family = "Calibri"), 
        legend.title=element_blank(),
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank(),
        element_line(linewidth = 1, colour = "black")
                )
```

```{r message=FALSE, warning=FALSE}
a <- ggplot(abiotic, aes(x = KF.soil.temp, y = flux.med, colour = month)) +
  geom_smooth(method = "lm", colour = "black") +
  geom_point( size = 3) +
  scale_colour_viridis_d(option = "magma", begin = 0.2, end = 0.8) +
  scale_x_continuous(n.breaks = 10) +
  my_theme +
  theme(axis.text.x = element_text(hjust = 1, vjust = 0.4)) +
  labs(x="Soil temperature", y="Median methane flux")

b <- ggplot(abiotic, aes(x = KF.plant.cover, y = flux.med, colour = month)) +
  geom_smooth(method = "lm", colour = "black") +
  geom_point( size = 3) +
  scale_colour_viridis_d(option = "magma", begin = 0.2, end = 0.8) +
  scale_x_continuous(n.breaks = 6) +
  my_theme +
  theme(axis.text.x = element_text(hjust = 1, vjust = 0.4)) +
  labs(x="Relative plant height", y="")

c <- ggplot(abiotic, aes(x = KF.air.temp, y = flux.med, colour = month)) +
  geom_smooth(method = "lm", colour = "black") +
  geom_point( size = 3) +
  scale_colour_viridis_d(option = "magma", begin = 0.2, end = 0.8) +
  scale_x_continuous(n.breaks = 6) +
  my_theme +
  theme(axis.text.x = element_text(hjust = 1, vjust = 0.4)) +
  labs(x="Air temperature", y="")

d <- ggplot(abiotic, aes(x = KF.VWC, y = flux.med, colour = month)) +
  geom_smooth(method = "lm", colour = "black") +
  geom_point( size = 3) +
  scale_colour_viridis_d(option = "magma", begin = 0.2, end = 0.8) +
  scale_x_continuous(n.breaks = 6) +
  my_theme +
  theme(axis.text.x = element_text(hjust = 1, vjust = 0.4)) +
  labs(x="Volumetric water content", y="")

legend <- get_legend(a)

grid <- plot_grid(a + theme(legend.position = 'none'),
          b + theme(legend.position = 'none'),
          c + theme(legend.position = 'none'),
          d + theme(legend.position = 'none'),
        legend, rel_heights = c(1,1,1,1,0.5), nrow = 2, ncol = 3)

grid

# split up plots and legend
plots <- plot_grid(a + theme(legend.position = 'none'),
          b + theme(legend.position = 'none'),
          c + theme(legend.position = 'none'),
          d + theme(legend.position = 'none'),
          rel_heights = c(1,1,1,1), nrow = 2, ncol = 2)

plots1 <- plot_grid(plots + theme(legend.position = 'none'),
          legend, rel_heights = c(1,0.2),rel_widths = c(1, 0.2) , nrow = 1, ncol = 2)

plots1

# put plots in one row
plot_grid(a + theme(legend.position = 'none'),
          b + theme(legend.position = 'none'),
          c + theme(legend.position = 'none'),
          d + theme(legend.position = 'none'),
          legend,
          rel_heights = c(.1,.1,.1,.1,0.1),
          rel_widths = c(.1,.1,.1,.1,0.05),
          nrow = 1, ncol = 5)

```

```{r}
L <- ggscatter(abiotic, x = "KF.air.temp", y = "flux.med",
          add = "reg.line",  
          conf.int = TRUE,                                       
          add.params = list(color = "black", fill = "lightgray"),
          cor.coef = TRUE, cor.method = "pearson",
          cor.coeff.args = list(label.x = 16, label.y = 20),
          repel = T) + 
        theme_bw() +
        theme(text=element_text(size = 10, family = "Arial"),
              axis.text.x=element_text(size=10, color = "black", family = "Arial"), 
              axis.text.y = element_text(size=10, color = "black",  family = "Arial"),
              title = element_text(size = 10, color = "black",  family = "Arial"), 
              legend.title=element_blank(),
              panel.grid.minor = element_blank(), 
              panel.grid.major = element_blank(),
              element_line(linewidth = 1, colour = "black")) +
  xlab("Air temperature (\u00b0C)") +
  ylab(expression(Methane ~ flux ~ (mu*g ~ CH[4]-C ~ ~ m^{-2} ~ hr^{-1})))

M <- ggscatter(abiotic, x = "KF.plant.cover", y = "flux.med",
          xlab = F,
          ylab = F,
          repel = T,
          add = "reg.line",
          conf.int = TRUE,  
          add.params = list(color = "black", fill = "lightgray"),
          cor.coef = TRUE, cor.method = "pearson",
          cor.coeff.args = list(label.x = 1, label.y = 20)) +
  theme_bw() +
        theme(text=element_text(size = 10, family = "Arial"),
              axis.text.x=element_text(size=10, color = "black", family = "Arial"), 
              axis.text.y = element_text(size=10, color = "black",  family = "Arial"),
              title = element_text(size = 10, color = "black",  family = "Arial"), 
              legend.title=element_blank(),
              panel.grid.minor = element_blank(), 
              panel.grid.major = element_blank(),
              element_line(linewidth = 1, colour = "black")) +
  xlab("Relative plant cover (cm)")+
  ylab("")

N <- ggscatter(abiotic, x = "KF.soil.temp", y = "flux.med",
          xlab = "air temperature (\u00b0C)",
          ylab = "Methane flux (\u03BCg CH4-C m-2 hr-1)",
          add = "reg.line",                                      
          conf.int = TRUE,  
          repel = T,
          add.params = list(color = "black", fill = "lightgray"),
          cor.coef = TRUE, cor.method = "pearson",
          cor.coeff.args = list(label.x = 5, label.y = 20)) +
   theme_bw() +
        theme(text=element_text(size = 10, family = "Arial"),
              axis.text.x=element_text(size=10, color = "black", family = "Arial"), 
              axis.text.y = element_text(size=10, color = "black",  family = "Arial"),
              title = element_text(size = 10, color = "black",  family = "Arial"), 
              legend.title=element_blank(),
              panel.grid.minor = element_blank(), 
              panel.grid.major = element_blank(),
              element_line(linewidth = 1, colour = "black")) +
  xlab("Soil temperature (\u00b0C)")+
  ylab("")

O <- ggscatter(abiotic, x = "KF.VWC", y = "flux.med",
          xlab = F,
          ylab = F,
          add = "reg.line",                                      
          conf.int = TRUE,  
          repel = T,
          add.params = list(color = "black", fill = "lightgray"),
          cor.coef = TRUE, cor.method = "pearson",
          cor.coeff.args = list(label.x = 12, label.y = 20)) +
     theme_bw() +
        theme(text=element_text(size = 10, family = "Arial"),
              axis.text.x=element_text(size=10, color = "black", family = "Arial"), 
              axis.text.y = element_text(size=10, color = "black",  family = "Arial"),
              title = element_text(size = 10, color = "black",  family = "Arial"), 
              legend.title=element_blank(),
              panel.grid.minor = element_blank(), 
              panel.grid.major = element_blank(),
              element_line(linewidth = 1, colour = "black")) +
  xlab("VWC") +
  ylab("")

KF_corr_plot <- plot_grid(L,M,N,O, nrow = 1, ncol = 4)
```

Methane flux

```{r}
# load in data
flux1 <- read.table("flux_data.csv", header = T, row.names = 1, sep = ",")

# make sure months don't get re-organized alphabetically
month_order <- flux1 %>%
  mutate(month = factor(month, levels = c("May", "June", "July", "August", "October")))

# plot
CH4_plot <- ggplot(data = month_order, aes(x=month, y=flux)) + 
    geom_boxplot(aes(fill = soil), show.legend = F) +
    scale_fill_manual(values = c("white", "white"))+
  my_theme +
  geom_point(aes(color=factor(soil)), position = position_dodge(width = 0.75), size=2.5) +
  scale_color_manual(values = c("#0073e6","#5ba300")) +
  xlab("") +
  ylab(expression(Methane ~ flux ~ (mu*g ~ CH[4]-C ~ ~ m^{-2} ~ hr^{-1}))) +
  geom_hline(yintercept = c(0)) +
  geom_vline(xintercept = c(0.4)) +
  theme(panel.border = element_blank(),
        axis.ticks.x = element_blank(), 
        legend.text = element_text(size = 10), 
        axis.text.y = element_text(size = 10), 
        legend.position = c(.17,.98), 
        legend.key.size = unit(.8, "line"), 
        plot.margin = margin(t = 20, r = 20, b = 5, l = 5))

print(CH4_plot)
```
 
Now we will look at the data that was collected at the site of the methane flux chambers, including the soil temperature measured by Puri lab members, rather than the remote research stations. The main difference is that we can include all of our replicate measurements here, whereas earlier we could only look at median values.

```{r}
# load in data
KF_df <- read.table("KF_data_for_plot.csv", header = T, row.names = 1, sep = ",")

# plot
KF_plot <- ggplot(KF_df, aes(x=month, y=mean, fill=enviro)) +
  geom_bar(width= 0.7, position=position_dodge(), stat="identity", colour='black') +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=.4,position=position_dodge(.7)) +
  scale_fill_manual(name = "enviro", values = c("#ff0000", "darkgray", "#fff800", "#00f0e5"), labels = c("air temperature (°C)", "relative plant height (cm)","soil temperature (°C)", "VWC" )) +
  my_theme +
  labs(x = "", y = "Average value") +
  scale_x_discrete(labels=c("May","June","July", "August","October")) +
  geom_hline(yintercept = c(0)) +
  geom_vline(xintercept = c(0.4)) +
  scale_y_continuous(breaks = c(-20, 0, 20, 40, 60, 80, 100, 120), limits = c(-20, 120)) +
   theme(panel.border = element_blank(),
        axis.ticks.x = element_blank(), 
        legend.text = element_text(size = 10), 
        axis.text.y = element_text(size = 10), 
        legend.position = c(.88,.98), 
        legend.key.size = unit(.8, "line"), 
        plot.margin = margin(t = 20, r = 43, b = 5, l = 20)) +
  guides(fill = guide_legend(# title.hjust = 1, # adjust title if needed
                             label.position = "left",
                             label.hjust = 1))

print(KF_plot)
```

Plot all three plots together
```{r}
# combine methane and enviro plots
top_plot <- plot_grid(CH4_plot, KF_plot, nrow = 1, rel_widths = c(1,1))

# combine with correlation plots
plot_grid(top_plot, KF_corr_plot, nrow = 2)

plot_grid(
  plot_grid(CH4_plot, KF_plot, nrow = 1, ncol = 2),
  plot_grid(NULL, KF_corr_plot, NULL, nrow = 1, rel_widths = c(0.01, 1, .01)),
  nrow = 2
)
```
Judd depth from each day at each time point from the year 2021
```{r}
judd_depth <- read.csv("judd_depth.csv", header = T, row.names = 1, sep = ",")

# Convert the date column in the data frame to POSIXct
judd_depth$date <- as.POSIXct(judd_depth$date, format = "%Y/%m/%d %H:%M")

# theme 
my_theme_grid1 <- theme_bw() +
        theme(text=element_text(size = 12), 
        axis.text.x=element_text(angle = 45, hjust=1, size=10, color = "black"), 
        axis.text.y = element_text(size=10, color = "black"),
        title = element_text(size = 10, color = "black"), 
        legend.title=element_blank(),
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank()
        )

# Example of plotting after conversion
ggplot(judd_depth, aes(x = date, y = JuddDepth_Avg)) +
  geom_line() +
  scale_x_datetime(date_breaks = "1 month", date_labels = "%B") + # Format for month and day
  labs(title = "", x = "", y = "Relative plant and snow depth (cm)") +
  my_theme_grid1
```
Instead, do the average from each day
```{r}
library(dplyr)
library(lubridate)

# Create a new column for day and month
judd_avg <- judd_depth %>%
  mutate(dayMonth = format(date, "%m-%d"))  # Extracts month and day

# Calculate the average for each day of the year
averagePerDay <- judd_avg %>%
  group_by(dayMonth) %>%
  summarize(averageValue = mean(JuddDepth_Avg, na.rm = TRUE))

# View the new data frame
print(averagePerDay)


averagePerDay <- averagePerDay %>%
  mutate(dayMonth = as.Date(paste("2021", dayMonth), format="%Y %m-%d"))


# Define custom breaks for each month
month_breaks <- seq(as.Date("2021-01-01"), as.Date("2021-12-31"), by = "month")

# Plotting
ggplot(averagePerDay, aes(x = dayMonth, y = averageValue)) +
  geom_line() +
  scale_x_date(breaks = month_breaks, labels = format(month_breaks, "%B-%d")) +
  scale_y_continuous(limits = c(-10,150)) +
  labs(title = "", x = "2021", y = "Relative plant and snow depth (cm)") +
  my_theme_grid1
```

volumetric water content at 5 cm from all time points
```{r}
vwc_daily <- read.csv("vwc_5cm.csv", header = T, row.names = 1, sep = ",")

# Convert the date column in the data frame to POSIXct
vwc_daily$date <- as.POSIXct(vwc_daily$date, format = "%Y/%m/%d %H:%M")

# theme 
my_theme_grid1 <- theme_bw() +
        theme(text=element_text(size = 12), 
        axis.text.x=element_text(angle = 45, hjust=1, size=10, color = "black"), 
        axis.text.y = element_text(size=10, color = "black"),
        title = element_text(size = 10, color = "black"), 
        legend.title=element_blank(),
        panel.grid.minor = element_blank(), 
        )

# Example of plotting after conversion
ggplot(vwc_daily, aes(x = date, y = VWC_5cm_Avg)) +
  geom_line() +
  scale_x_datetime(date_breaks = "1 month", date_labels = "%B") + # Format for month and day
  labs(title = "", x = "", y = "volumetric water content at 5 cm depth") +
  my_theme_grid1


# Create a new column for day and month
vwc_avg <- vwc_daily %>%
  mutate(dayMonth = format(date, "%m-%d"))  # Extracts month and day

# Calculate the average for each day of the year
averagePerDay_vwc <- vwc_avg %>%
  group_by(dayMonth) %>%
  summarize(averageValue = mean(VWC_5cm_Avg, na.rm = TRUE))

# View the new data frame
print(averagePerDay_vwc)


averagePerDay_vwc <- averagePerDay_vwc %>%
  mutate(dayMonth = as.Date(paste("2021", dayMonth), format="%Y %m-%d"))


# Define custom breaks for each month
month_breaks <- seq(as.Date("2021-01-01"), as.Date("2021-12-31"), by = "month")

# Plotting
ggplot(averagePerDay_vwc, aes(x = dayMonth, y = averageValue)) +
  geom_line() +
  scale_x_date(breaks = month_breaks, labels = format(month_breaks, "%B-%d")) +
  scale_y_continuous(limits = c(0,40)) +
  labs(title = "", x = "2021", y = "volumetric water content at 5 cm depth") +
  my_theme_grid1

```

```{r}
# stack these on top with a shared x axis

judd_plot <- ggplot(averagePerDay, aes(x = dayMonth, y = averageValue)) +
  geom_line() +
  scale_x_date(breaks = month_breaks) +
  scale_y_continuous(limits = c(-10,150)) +
  labs(title = "", x = "", y = "Relative plant and snow depth (cm)") +
  my_theme_grid1 +
  theme(axis.text.x = element_blank())

vwc_plot <- ggplot(averagePerDay_vwc, aes(x = dayMonth, y = averageValue)) +
  geom_line() +
  scale_x_date(breaks = month_breaks, labels = format(month_breaks, "%B-%d")) +
  scale_y_continuous(limits = c(0,40)) +
  labs(title = "", x = "2021", y = "VWC at 5 cm depth") +
  my_theme_grid1 

plot_grid(judd_plot, vwc_plot, nrow = 2)
```

