---
title: "Heatmap"
output: html_notebook
---

Following along with How to use R to create a heatmap from a distance matrix with ggplot2 (CC209) by Riffomonas Project (<https://www.youtube.com/watch?v=3iOTGaWQZp4>)

```{r}
library(tidyverse)
library(vegan)
```

```{r}
gDNA_comm_tbl <- bact_count_table_t %>%
    as_tibble(rownames = "sample") %>%
    pivot_longer(-sample) %>%
    inner_join(., sample_lookup, by="sample") %>%
    filter(template %in% gDNA_wanted) %>%
    group_by(sample) %>%
    mutate(N = sum(value)) %>%
    ungroup() %>%
    select(-N) %>%
    pivot_wider(names_from = "name", values_from = "value", values_fill = 0) %>%
    select(-month, -soil, -template, -replicate, -names) %>%
    column_to_rownames("sample") 

head(gDNA_comm_tbl)
```

In this step we calculate the avg distance which creates a distance matrix, but if we want to do a heatmap we need to turn the distance matrix into a matrix and then that matrix into a tibble. We will temporarily call the row names "A" because we will take the column names and make those another column name for "B"

```{r}
gDNA_comm_dist <- avgdist(gDNA_comm, dmethod="bray", sample = 116838) %>%
  as.matrix() %>%
  as_tibble(rownames="A")

gDNA_comm_dist
```

That gives a square tibble. Now if we want to take the column names and make this longer:

```{r}
bray <- gDNA_comm_dist %>%
  pivot_longer(-A, names_to="B", values_to = "distances")

bray
```

Great, now we have our tibble in long form.

Let's plot bray.

```{r}
bray %>%
  ggplot(aes(x=A, y=B, fill=distances)) +
  geom_tile()
```

now lets look at a heatmap made with jaccard instead of bray

```{r}
jaccard <- avgdist(gDNA_comm, dmethod="jaccard", sample = 116838) %>%
  as.matrix() %>%
  as_tibble(rownames="A")

jaccard2 <- gDNA_comm_dist %>%
  pivot_longer(-A, names_to="B", values_to = "distances")

jaccard2 %>%
  ggplot(aes(x=A, y=B, fill=distances)) +
  geom_tile()
```

no difference!except maybe jaccard distances are more different since the colors are a bit lighter
