---
title: "RBC 16S analysis: MOB visualizations (family level)"
output: github_document
---
Written by Delaney Beals \| Date created: January 23, 2024 \| Date most recent edit: January 23, 2024

# Relative abundance composition plot for family-level proportions

Resource: <https://astrobiomike.github.io/amplicon/dada2_workflow_ex>

```{r message=FALSE, warning=FALSE}
# load in libraries
library(tidyverse) 
library(phyloseq)
library(vegan)
library(DESeq2)
library(dendextend)
library(viridis)
library(ggplot2)
library(readr)
library(gridExtra)
library(cowplot)
library(dplyr)
library(forcats)
library(patchwork)
library(ggpubr)
library(corrplot)
(.packages())
```

Making our phyloseq object
```{r}
# LOAD IN DATA
bact_count_table <- read.table("June_Oct_MERGE_ASV_counts.tsv", header = T, row.names = 1, check.names = F, sep = "\t") 
bact_taxa_table <- as.matrix(read.table("June_Oct_MERGE_ASV_taxa_table.tsv", header = T, row.names = 1, check.names = F, sep = "\t"))
bact_sample_info_table <- read.table("June_Oct_16S_MERGE_metadata - Copy.csv", header = T, row.names = 1, sep = ",") 

# first we need to create a phyloseq object using our un-transformed count table
tax_mat <- as.matrix(bact_taxa_table)
TAX <- phyloseq::tax_table(tax_mat)

samples = sample_data(bact_sample_info_table)

otu_mat <- as.matrix(bact_count_table)
OTU = otu_table(otu_mat, taxa_are_rows = TRUE)
ASV_physeq <- phyloseq(TAX, OTU, samples)
```

## Calculating relative abundance

First make a table with proportions of reads. This will essentially be a normalization step in that we are calculating relative abundance of every ASV within each sample.
```{r}
fam_counts_tab <- otu_table(tax_glom(ASV_physeq, taxrank="Family")) 

family_tax_vec <- as.vector(phyloseq::tax_table(tax_glom(ASV_physeq, taxrank="Family"))[,5]) 

rownames(fam_counts_tab) <- as.vector(family_tax_vec)

unclassified_tax_counts_fam <- colSums(bact_count_table) - colSums(fam_counts_tab)

# and we'll add this row to our phylum count table:

fam_and_unidentified_counts_tab <- rbind(fam_counts_tab, "Unclassified"=unclassified_tax_counts_fam)

identical(colSums(fam_and_unidentified_counts_tab), colSums(bact_count_table)) # TRUE

fam_taxa_proportions_tab <- apply(fam_and_unidentified_counts_tab, 2, function(x) x/sum(x)*100)

#transform the table so that genera are in columns

write.table(fam_taxa_proportions_tab, "fam_proportions.tsv", sep = "\t", quote = F, col.names = NA)
```


```{r}
# transpose table so that samples are in rows, genera are in columns
fam_taxa_proportions_tab_df <- as.data.frame(t(fam_taxa_proportions_tab))

# classify our families into functional groups
methanotrophs_fam <- c("Methylocystaceae|Beijerinckiaceae|Methylococcaceae|Methylothermaceae|Crenotrichaceae|Methylacidiphilaceae")

methylotrophs_fam <- c("Methylocystaceae|Methylobacteriaceae|Hyphomicrobiaceae|Rhodobacteraceae|Comamonadaceae|Oxalobacteraceae|Rhodocyclaceae|Methylophilaceae|Thiotrichaceae|Flavobacteriaceae|Burkholderiaceae")

meth_all_fam <- c("Methylocystaceae|Beijerinckiaceae|Methylococcaceae|Methylothermaceae|Crenotrichaceae|Methylacidiphilaceae|Methylocystaceae|Methylobacteriaceae|Hyphomicrobiaceae|Rhodobacteraceae|Comamonadaceae|Oxalobacteraceae|Rhodocyclaceae|Methylophilaceae|Thiotrichaceae|Flavobacteriaceae|Burkholderiaceae")

# create new tables containing only the genera groups we want
methanotrophs_prop_fam <- fam_taxa_proportions_tab_df %>% select(matches(methanotrophs_fam))
methylotrophs_prop_fam <- fam_taxa_proportions_tab_df %>% select(matches(methylotrophs_fam))
meth_all_prop_fam <- fam_taxa_proportions_tab_df %>% select(matches(meth_all_fam))

# transpose table again so that rows are genera and columns are samples
methanotrophs_prop1_fam <- as.data.frame(t(methanotrophs_prop_fam))
methylotrophs_prop1_fam <- as.data.frame(t(methylotrophs_prop_fam))
meth_all_prop1_fam <- as.data.frame(t(meth_all_prop_fam))

# make sure we are in a data frame
methanotrophs_prop_tab_fam <- methanotrophs_prop1_fam %>%
  as_tibble(rownames= "Family") %>% 
  as.data.frame(rownames=paste0("Family", 1:24)) 

methylotrophs_prop_tab_fam <- methylotrophs_prop1_fam %>%
  as_tibble(rownames= "Family") %>% 
  as.data.frame(rownames=paste0("Family", 1:24))

meth_all_prop_tab_fam <- meth_all_prop1_fam %>%
  as_tibble(rownames= "Family") %>% 
  as.data.frame(rownames=paste0("Family", 1:24))

MOB_prop_fam <- pivot_longer(methanotrophs_prop_tab_fam, !Family, names_to = "Sample", values_to = "Proportion") %>% data.frame()
NMM_prop_fam <- pivot_longer(methylotrophs_prop_tab_fam, !Family, names_to = "Sample", values_to = "Proportion") %>% data.frame()
prop_fam <- pivot_longer(meth_all_prop_tab_fam, !Family, names_to = "Sample", values_to = "Proportion") %>% data.frame()

# merging this table with the table containing our proportions
prop_info_fam <- merge(prop_fam, sample_info_for_merge) 
MOB_prop_info_fam <- merge(MOB_prop_fam, sample_info_for_merge) 
NMM_prop_info_fam <- merge(NMM_prop_fam, sample_info_for_merge) 
```

```{r}
# set the ggplot2 theme
my_theme_grid <- theme_bw() +
        theme(text=element_text(size = 12), 
        axis.text.x=element_text(size=10, color = "black"), 
        axis.text.y = element_text(size=10, color = "black"),
        title = element_text(size = 10, color = "black"), 
        legend.title=element_blank()
        )
```


```{r}
# reordering the labels
new_prop_info_fam <- prop_info_fam %>%
  mutate(Family = fct_relevel(Family,  "Flavobacteriaceae", "Comamonadaceae", "Burkholderiaceae","Rhodocyclaceae","Rhodobacteraceae","Oxalobacteraceae","Hyphomicrobiaceae","Methylocystaceae","Beijerinckiaceae","Methylococcaceae","Methylobacteriaceae"))

# overall plot
ggplot(new_prop_info_fam, aes(Proportion, Family)) +
  geom_jitter(aes(color=factor(monthsoil), shape=factor(template)), 
              size=2.5, width=0.15, height=0) +
  scale_x_continuous(breaks = c(0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15)) +
  my_theme_grid

# zooming into smaller percentages
ggplot(new_prop_info_fam, aes(Proportion, Family)) +
  geom_jitter(aes(color=factor(monthsoil), shape=factor(template)), 
              size=2.5, width=0.15, height=0) +
  xlim(0, 0.5)+
  my_theme_grid 
```
```{r}
prop_info_gDNA_fam <-meth_all_prop_fam[grepl("gDNA", rownames(meth_all_prop_fam)), ]
prop_info_cDNA_fam <-meth_all_prop_fam[grepl("cDNA", rownames(meth_all_prop_fam)), ]
```

```{r}
# re-order families
desired_order <- c("Beijerinckiaceae","Methylobacteriaceae", "Methylococcaceae", "Methylocystaceae","Methylophilaceae", 
                   "Flavobacteriaceae", "Comamonadaceae", "Burkholderiaceae") 

# Reorder the columns of the dataframe
new_com_gDNA_fam <- prop_info_gDNA_fam[, desired_order]

# create a correlation matrix with the data
cc_pears_fam = cor(new_com_gDNA_fam, method = "pearson")
```

```{r}
# to get P values
res2 <- cor.mtest(new_com_gDNA_fam, conf.level = .95)

#merged option
corrplot.mixed(cc_pears_fam, upper= "circle",   tl.pos = "lt", tl.col = "black", tl.cex = 0.7, p.mat = res2$p, insig = "pch", sig.level = c(.05), pch.cex = 2, pch.col = "black", lower = "number",  number.cex = 0.7, tl.srt = 45)
```
```{r}
# re-order families
desired_order <- c("Beijerinckiaceae","Methylobacteriaceae", "Methylococcaceae", "Methylocystaceae","Methylophilaceae", 
                   "Flavobacteriaceae", "Comamonadaceae", "Burkholderiaceae") 

# Reorder the columns of the dataframe
new_com_cDNA_fam <- prop_info_cDNA_fam[, desired_order]

# create a correlation matrix with the data
cc_pears_fam_c = cor(new_com_cDNA_fam, method = "pearson")
```

```{r}
# to get P values
res3 <- cor.mtest(new_com_cDNA_fam, conf.level = .95)

#merged option
corrplot.mixed(cc_pears_fam_c, upper= "circle",   tl.pos = "lt", tl.col = "black", tl.cex = 0.7, p.mat = res3$p, insig = "pch", sig.level = c(.05), pch.cex = 2, pch.col = "black", lower = "number",  number.cex = 0.7, tl.srt = 45)
```

```{r}
MOB_NMH_gDNA_fam <- read.table("MOB_NMH_gDNA_fam.csv", header = T, row.names = 1, sep = ",")
```

```{r}
#calculate correlation
gf_df <- Hmisc::rcorr(as.matrix(MOB_NMH_gDNA_fam), type=c("pearson"))

corrplot(corr=gf_df$r[13:17, 1:12], 
         p.mat=gf_df$P[13:17, 1:12], 
         tl.col = "black",
         tl.srt = 45,
         insig = "pch", 
         sig.level = c(.05), 
         pch.cex = 3, 
         pch.col = "black", 
         addCoef.col="black",  
         number.cex=0.7,
         number.digits = 2,
         cl.length = 5,
         cl.align.text = "l")

```

```{r}
MOB_NMH_cDNA_fam <- read.table("MOB_NMH_cDNA_fam.csv", header = T, row.names = 1, sep = ",")
```

```{r}
#calculate correlation
cf_df <- Hmisc::rcorr(as.matrix(MOB_NMH_cDNA_fam), type=c("pearson"))

corrplot(corr=cf_df$r[13:17, 1:12], 
         p.mat=cf_df$P[13:17, 1:12], 
         tl.col = "black",
         tl.srt = 45,
         insig = "pch", 
         sig.level = c(.05), 
         pch.cex = 3, 
         pch.col = "black", 
         addCoef.col="black",  
         number.cex=0.7,
         number.digits = 2,
         cl.length = 9,
         cl.align.text = "l")
```
```

