---
title: "MICROBIOME SIGNIFICANCE"
output: html_notebook
---

source: https://wiki.duke.edu/display/AnthroTree/11.1+Phylogenetic+community+ecology+in+R

```{r}
library(picante)
```

We want our data to be in the format of community ID in row headers, ASVs in column headers, and proportions in cells using bact_count_table, calculate the proportions

```{r}
ASV_proportions_tab <- apply(bact_count_table, 2, function(x) x/sum(x)*100)
head(ASV_proportions_tab)
```

transform  our data to be in the format of community ID in row headers, ASVs in column headers, and proportions in cells
```{r}
ASV_proportions_tab_t <- as.data.frame(t(ASV_proportions_tab))
head(ASV_proportions_tab_t)
```

Create our phylogeny, which is contained within our ASV_phyloseq that we made in earlier analysis steps
```{r}
ASV_tree = rtree(ntaxa(ASV_physeq), rooted=TRUE, tip.label=taxa_names(ASV_physeq))
str(ASV_tree)
```


picante assumes that the taxa in your phylogeny and the taxa in your dataset are in the SAME ORDER. This is easy to do:
```{r}
ASV_proportions_tab_t<-ASV_proportions_tab_t[,ASV_tree$tip.label]

head(ASV_proportions_tab_t)
```



We can now look at the distribution of species within our three communities across the phylogeny. Really big, this takes a while to run and is honestly impossible to read. 

```{r}
par(mfrow = c(2, 2))
plot(ASV_tree)
for (i in rownames(ASV_proportions_tab_t)) {
  plot(ASV_tree, show.tip.label = FALSE, main = i)
  tiplabels(tip = which(ASV_proportions_tab_t[i, ] > 0), pch = 19, cex = 2, col ="red")
  legend("topleft" , i, bty = "n")
}
```


## COMMUNITY PHYLOGENETICS MEASURES

Phylogenetic diversity (Faith 1992): Phylogenetic diversity (PD) calculates the total branch length spanned by the species within a community. To calculate PD for our communities:
```{r}
bact.pd <- pd(ASV_proportions_tab_t, ASV_tree, include.root = TRUE)
bact.pd
```
This includes all our data in the PD calculation. Let's separate the data by template type and then run pd function again to see how the numbers change. 

```{r}
# create a vector so we only select gDNA-derived samples
gDNA_wanted <- c("gDNA")

ASV_df <- ASV_proportions_tab_t %>%
    as_tibble(rownames = "sample") %>%
    pivot_longer(-sample) %>%
    inner_join(., sample_lookup, by="sample") %>%
    filter(template %in% gDNA_wanted) %>%
    select(-month, -soil, -template, -replicate, -names, -flux) %>%
    group_by(sample) %>%
    pivot_wider() %>%
    as.data.frame(rownames=paste0("sample", 1:12)) %>%
    column_to_rownames("sample")

head(ASV_df)
```

```{r}
gDNA.pd <- pd(ASV_df, ASV_tree, include.root = TRUE)
gDNA.pd
```

Okay so these values are the same as when we kept the data set all together, which means we don't have to separate by template type and run in parallel. 


```{r}
bact.pd.tbl <- bact.pd %>%
  as_tibble(rownames="sample")
ASV_proportions_tbl <- ASV_proportions_tab_t %>%
  as_tibble(rownames="sample")


ASV_info <- ASV_proportions_tab_t %>%
    as_tibble(rownames = "sample") %>%
    inner_join(., sample_lookup, by="sample") %>%
    select(sample, month, soil, template, replicate, flux) %>%
    group_by(sample) %>%
    inner_join(., bact.pd.tbl, by="sample") %>%
    as.data.frame(rownames=paste0("sample", 1:24)) %>%
    column_to_rownames("sample")

ASV_info
```

Now calculate distance matrices
```{r}
ASV_euc_dist <- dist(ASV_info)

ASV_euc_clust <- hclust(ASV_euc_dist, method="ward.D2")

plot(ASV_euc_clust)
```


Remove non-integer values like month, soil type. 
```{r}

ASV_PD <- ASV_info %>%
    as_tibble(rownames = "sample") %>%
    select(sample, flux, PD, SR) %>%
    group_by(sample) %>%
    as.data.frame(rownames=paste0("sample", 1:24)) %>%
    column_to_rownames("sample")

PD_euc_dist <- dist(ASV_PD)

PD_euc_clust <- hclust(PD_euc_dist, method="ward.D2")
plot(PD_euc_clust)
```

