---
title: 'RBC 16S analysis: UpSet plots'
output: html_document
---

Written by Delaney Beals \| Date created: October 23, 2023 \| Date most recent edit: December 11, 2023

# Shared and unique ASVs in samples

The goal is to look at what ASVs are unique to each sample and which ASVs are shared between different sets of samples. This analysis does not take into account abundances of ASVs, only if a given ASV is detected or not detected.

```{r}
library(phyloseq)
library(MicrobiotaProcess)
library(UpSetR)
```

```{r}
# load in data
bact_count_table <- read.table("June_Oct_MERGE_ASV_counts.tsv", header = T, row.names = 1, check.names = F, sep = "\t") 
bact_taxa_table <- as.matrix(read.table("June_Oct_MERGE_ASV_taxa_table.tsv", header = T, row.names = 1, check.names = F, sep = "\t"))
bact_sample_info_table <- read.table("June_Oct_16S_MERGE_metadata.csv", header = T, row.names = 1, sep = ",")

# transform to phyloseq objects
OTU = phyloseq::otu_table(bact_count_table, taxa_are_rows = TRUE)
TAX = phyloseq::tax_table(bact_taxa_table)
samples = sample_data(bact_sample_info_table)

# create phyloseq object
ASV_physeq <- phyloseq(OTU, TAX, samples)

# generate data for upset plot
upset.ASV <- get_upset(obj = ASV_physeq, factorNames = "replicate")
```

Generate an UpSet plot showing all samples.

```{r}
upset(upset.ASV, sets=unique(as.vector(sample_data(ASV_physeq)$replicate)),
      sets.bar.color = "#56B4E9")

upset(upset.ASV, sets=unique(as.vector(sample_data(ASV_physeq)$replicate)),
      order.by = "degree",
      sets.bar.color = "#56B4E9")
```

Breaking up into more plots that are easier to view.

```{r}
upset.ASV %>% upset(., keep.order = T, sets = c("June-rip-gDNA", "June-up-gDNA", "Oct-rip-gDNA", "Oct-up-gDNA"), sets.bar.color = "#56B4E9", order.by = "degree") 

upset.ASV %>% upset(., keep.order = T, sets = c("June-rip-cDNA", "June-up-cDNA", "Oct-rip-cDNA", "Oct-up-cDNA"), sets.bar.color = "#56B4E9", order.by = "degree")

upset.ASV %>% upset(., keep.order = T, sets = c("June-rip-gDNA", "June-rip-cDNA", "June-up-gDNA", "June-up-cDNA"), sets.bar.color = "#56B4E9", order.by = "degree")

upset.ASV %>% upset(., keep.order = T, sets = c("Oct-rip-gDNA", "Oct-rip-cDNA", "Oct-up-gDNA", "Oct-up-cDNA"), sets.bar.color = "#56B4E9", order.by = "degree")
```

```{r}
upset(upset.ASV, 
      keep.order = T, 
      sets = c("June-rip-gDNA","June-up-gDNA",
               "Oct-rip-gDNA","Oct-up-gDNA"),
      sets.bar.color = "#56B4E9", 
      order.by = "degree",
      sets.x.label = "total ASV count", 
      mainbar.y.label = "ASV intersection size", 
      text.scale = c(1.5, 1.5, 1.25, 1, 1.25, 1.25))

                
                
upset(upset.ASV, 
      keep.order = T, 
      sets = c("June-rip-cDNA","June-up-cDNA",
               "Oct-rip-cDNA","Oct-up-cDNA"), 
      sets.bar.color = "#56B4E9", 
      order.by = "degree",
      sets.x.label = "total ASV count", 
      mainbar.y.label = "ASV intersection size", 
      text.scale = c(1.5, 1.5, 1.25, 1, 1.25, 1.25))
```
